---
title: "Assignment 2: Coding Basics"
author: Logan Loadholtz
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your first and last name into the file name (e.g., "FirstLast_A02_CodingBasics.Rmd") prior to submission.



## Basics Day 1

1. Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name. 

2. Compute the mean and median of this sequence. 

3. Ask R to determine whether the mean is greater than the median. 

4. Insert comments in your code to describe what you are doing.

```{r}
#1. 
mysequence <- seq(1,100,4)
mysequence
#Here I first created a name for the sequence "mysequence". I assigned this name to the 
#sequence. I typed seq to tell R I am creating a sequence. I typed 1 to tell R where 
#I want the sequence to begin, and 100 to tell R where I want the sequence to end. 
#Finally, the 4 indicates how often to display values in the sequence, here indicates 
#increasing every fourth value.

#2. 
mymean <- mean(mysequence)
#Here, I am telling R to calculate the mean of the sequence, and I am assigning it 
#to the name mymean.

mymedian <- median(mysequence)
#Here, I am telling R to calculate the median of the sequence, and I am assigning 
#it to the name mymedian.

#3. 
mymean > mymedian
#Here, I am using a conditional statement where mean is greater than median. The output of 
#this says that this is FALSE" 
```

## Basics Day 2

5. Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6. Label each vector with a comment on what type of vector it is.

7. Combine each of the vectors into a data frame. Assign the data frame an informative name.

8. Label the columns of your data frame with informative titles.

```{r}
#5
studentnames <- c("Logan", "Carly", "Hannah", "Helen", "Elise", "Erik", "Tim", "Henry")
studentnames
typeof(studentnames)
#This indicates that this is a "character" vector.


testscore <- c(97, 82, 99, 93, 85, 100, 48, 36)
testscore
typeof(testscore)
#This indicates that this is a "double" vector


passingscores <- testscore > 50
passingscores
typeof(passingscores)
#This indicates that this is a "logical" vector.

studentstestgrades <- data.frame(studentnames, testscore, passingscores)
studentstestgrades


```

9. QUESTION: How is this data frame different from a matrix?

> Answer: A matrix containes one type of data. This dataframe can contain multiple types of data, such as character, double, and logical.

10. Create a function with an if/else statement. Your function should determine whether a test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. Hint: Use `print`, not `return`. The name of your function should be informative.

11. Apply your function to the vector with test scores that you created in number 5. 

```{r}

didstudentpassexam <- ifelse(testscore >50, TRUE, FALSE)
 
didstudentpassexam
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: I utilized the "ifelse" option. This worked because the output shows scores that are less than 50 as "FALSE". Scores above 50 show up as "TRUE". FOr this, I used the vector "testscore", which is the vector of exam scores. Then, I typed >50 to tell R that if there is an exam score, to mark it as "TRUE". Anything less than this will create an output of "FALSE".

