---
title: "Assignment 5: Data Visualization"
author: "Logan Loadholtz"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A05_DataVisualization.Rmd") prior to submission.

The completed exercise is due on Tuesday, February 23 at 11:59 pm.



## Set up your session 
1. Set up your session. Verify your working directory and load the tidyverse 
and cowplot packages. Upload the NTL-LTER processed data files for nutrients 
and chemistry/physics for Peter and Paul Lakes (both the tidy 
[`NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv`] and the gathered
[`NTL-LTER_Lake_Nutrients_PeterPaulGathered_Processed.csv`] versions) and 
the processed data file for the Niwot Ridge litter dataset. 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r}
#1 
getwd()
library(tidyverse)
install.packages("cowplot")
library(cowplot)



NTL_LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed <- read_csv("Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv")



NTL_LTER_Lake_Nutrients_PeterPaulGathered_Processed <- read_csv("Data/Processed/NTL-LTER_Lake_Nutrients_PeterPaulGathered_Processed.csv")


NEON_NIWO_Litter_mass_trap_Processed <- read_csv("Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv")


#2 

class(NTL_LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed$sampledate)
class(NTL_LTER_Lake_Nutrients_PeterPaulGathered_Processed$sampledate)
class(NEON_NIWO_Litter_mass_trap_Processed$collectDate)
#All three datasets are reading the dates as "date".

```
## Define your theme
3. Build a theme and set it as your default theme.

```{r}

MyDefaultTheme <- theme_classic(base_size = 13) +
  theme(axis.text = element_text(color = "blue"),
        legend.position = "top")

theme_set(MyDefaultTheme)

```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values.

```{r}
Plot1_Phos <- ggplot(NTL_LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed, aes(x= po4, y=tp_ug)) +
  geom_point() +
  geom_smooth(method = lm) +
  #xlim(0,46) +
  labs(title="Plot of Total Phosphorus by Phosphate",
  x="Phosphate",
  y="Total Phosphorus")
  

print(Plot1_Phos) 


```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned.

```{r}
#Boxplot1_temp <- ggplot(NTL_LTER_Lake_Nutrients_PeterPaul_Processed, aes(x= month, y=temp)) +
  geom_boxplot(aes(color="lakename")) +
  labs(title="Temperature by Month",
  x="Month",
  y="Temperature")

print(Boxplot1_temp)

#Boxplot2_TP <- ggplot(NTL_LTER_Lake_Nutrients_PeterPaul_Processed, aes(x= month, y=tp_ug) +
  geom_boxplot(aes(color="lakename")) +
  labs(title="Total Phosphorus by Month",
  x="Month",
  y="Total Phosphorus")

print(Boxplot2_TP)

#Boxplot3_TN <- ggplot(NTL_LTER_Lake_Nutrients_PeterPaul_Processed, aes(x= month, y=tn_ug)) +
  geom_boxplot(aes(color="lakename")) +
  labs(title="Total Nitrogen by Month",
  x="Month",
  y="Total Nitrogen")

print(Boxplot3_TN)

```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r}

Plot6_NiwotRidge <- ggplot(subset(NEON_NIWO_Litter_mass_trap_Processed, 
                                  functionalGroup == "Needles")) +
  geom_point(aes(x=collectDate, y=dryMass, color=nlcdClass))

print(Plot6_NiwotRidge)


Plot7_NiwotRidge <- ggplot(subset(NEON_NIWO_Litter_mass_trap_Processed, 
                                  functionalGroup == "Needles")) +
  geom_point(aes(x=collectDate, y=dryMass)) +
  facet_wrap(vars(nlcdClass), nrow=3)

print(Plot7_NiwotRidge)

```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: I think plot 7 is more effective becuase it displays the data more clearly in a way that is easier for the reader to comprehend. Plot 6 combines all three classes, and even though they are separated by color, it's still a little difficult to really see what is being displayed, especially if you want to analyze one class. Plot 7 is much more easy to read because the classes are separated, but even though they are separated, the years line up at the bottom for all 3 so you can still compare each one over time. Plot 7 allows the reader to easily analyze each class over time.
