---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Logan Loadholtz"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A06_GLMs.Rmd") prior to submission.

The completed exercise is due on Tuesday, March 2 at 1:00 pm.

## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup}
#1
getwd()
#install.packages("htmltools")
library(htmltools)
#install.packages("agricolae")
library(agricolae)
library(tidyverse)
library(lubridate)
library(ggplot2)

#importing data
NTL_LTER_Lake_Chemistry_Physics_Raw <- read.csv("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv", 
                                                stringsAsFactors = FALSE)

#find class of sample date, it is character
class(NTL_LTER_Lake_Chemistry_Physics_Raw$sampledate)

#format sampledate as DATE
NTL_LTER_Lake_Chemistry_Physics_Raw$sampledate <- as.Date(NTL_LTER_Lake_Chemistry_Physics_Raw$sampledate, 
                                                          format= "%m/%d/%y")

#sampledate is now in Date form
class(NTL_LTER_Lake_Chemistry_Physics_Raw$sampledate)

#2
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: Mean lake temperature recorded during July does not change with depth across all Lakes
Ha: Mean lake temperature recorded during July changes with depth across all lakes

4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4

NTL_Processed <- NTL_LTER_Lake_Chemistry_Physics_Raw %>%
  filter(daynum >181 & daynum <213) %>%
  select(lakename, year4, daynum, depth, temperature_C) %>%
  na.omit
  

#5
Plot_Temp_Depth <- ggplot(NTL_Processed, aes(x=depth, y=temperature_C)) +
  geom_point()+
  geom_smooth(method = lm, color="black")+
  ylim(0, 35)+
  xlab("Depth")+
  ylab("Temperature in Celcius")

print(Plot_Temp_Depth)

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: 
According to the line, there is a general negative trend between temperature and depth- the deeper the water, the lower the temperature is. However, with a closer look, this distribution shows that most of the points appear to be distributed towards lower depths, meaning that at lower depths, there can actually have quite a wide range of temperatures. As depth increases, temperature varies much less.

7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7

LTR_regression <- lm(NTL_Processed$temperature_C ~ NTL_Processed$depth)

summary(LTR_regression)

cor.test(NTL_Processed$temperature_C, NTL_Processed$depth)

par(mfrow = c(2,2), mar=c(2,2,2,2))
plot(LTR_regression)
par(mfrow = c(1,1))


```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: With a correlation coefficient of -0.859,  there is a strong negative correlation between temperature and depth (lower temperatures at greater depths). With an R-squared value of 0.7391, this model explains 73.9% of the total variance in temperature. This model has 9720 degrees of freedom, meaning a very large sample size. The result is statistically significant because the p value is 2.2e-16, which is much less than 0.05. For every 1m change in depth, temperature will decrease by 1.95 degrees Celcius.

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
NTL_AIC <- lm(data = NTL_Processed, temperature_C ~ year4 + daynum + depth)

step(NTL_AIC)


#10

NTLmultregression <- lm(data=NTL_Processed, temperature_C ~ year4 + daynum + depth)

summary(NTLmultregression)

par(mfrow = c(2,2), mar=c(4,4,4,4))
plot(NTLmultregression)
par(mfrow = c(1,1))

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: When running the AIC, it appears that all three variables should be used to suggest change in temperature. This is because the "Call" function shows that all three are included. 
-With an R squared value of 0.74, this explains that 74% of the variability of temperature can be explained by year, daynum, and depth. Each of the three p values is less than 0.05. 
-When using only depth as the explanatory variable, there R squared was 0.73. This means that there is only a slight improvement because ours increased to only 0.74

---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12

#ANOVA model
NTL_anova<- aov(data = NTL_Processed, temperature_C ~ lakename)

summary(NTL_anova)

plot(NTL_anova)

#Linear Model
NTL_anova2 <- lm(data = NTL_Processed, temperature_C ~ lakename)

summary(NTL_anova2)

plot(NTL_anova2)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Yes there is a significant difference in mean temperatures because the p value is less than 0.05.  

14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.

Plot_temp_lakes <- ggplot(NTL_Processed, aes(x=depth, y=temperature_C, color=lakename)) +
  geom_point(alpha=0.5)+
  geom_smooth(method="lm", se=FALSE) +
  ylim(0, 35)+
  xlab("Depth")+
  ylab("Temperature in Celcius")+
  theme(axis.text.x = element_text(angle=90))
  

print(Plot_temp_lakes)

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15

NTL_anova_3<- aov(data = NTL_Processed, temperature_C ~ lakename)

TukeyHSD(NTL_anova_3)


NTL_anova_comparison <- HSD.test(NTL_anova_3, "lakename", group = TRUE)
NTL_anova_comparison

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: The lakes that have the same mean temperature as Peter Lake are: Paul Lake and Ward Lake (according to the HSD.test). According to the HSD.test, all of the lakes are statistically related to at least one other lake. 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: 
If we were just comparing these two lakes, we could use a two-sample t-test. A two-sample t-test tests the hypothesis that the mean of two samples is equivalent.This would allow us to compare the means of each of the two lakes.

